---
title: "Accessible Course Document Template"
subtitle: "Title II ADA Compliant Configuration"
author: "Professor Name"
date: last-modified
lang: en                        # 1. Critical for screen reader pronunciation
format:
  html:
    theme: cosmo                # 2. Clean, high-contrast, accessible sans-serif typography
    toc: true                   # 3. Generates semantic visual & aria landmark navigation
    toc-depth: 3
    toc-location: left
    code-fold: true             # 4. Keeps cognitive load low; hides large code chunks
    code-tools: true            # Allows downloading code cleanly
    embed-resources: true       # Makes the HTML completely self-contained for distribution
    css: accessible-styles.css  # 5. Overrides default contrast & focus states
---

## Document Structure (Headings) {#sec-structure}

To comply with **ADA Title II (WCAG 2.1/2.2 Level AA)**, your document must follow a logical, nesting hierarchy. Screen readers rely entirely on heading tags to build a page index for navigation.

*   Always use structural headers sequentially (`## Heading 2`, `### Heading 3`).
*   **Never** skip a heading level (e.g., jumping from `## Heading 2` directly to `#### Heading 4`).
*   **Never** use bold text variations (`**Text**`) just to make a line look like a section title; screen readers will not recognize it as a page landmark.

## Computational Outputs & Tables {#sec-tables}

Do not print raw R console outputs (like basic matrix printouts) for tables, as screen readers cannot interpret them semantically. Use accessible layout packages like `knitr::kable` or `gt` to automatically generate semantic HTML structure (`<table>`, `<th>`, `<tr>`, `<td>`).

```{r}
#| label: tbl-demographics
#| tbl-cap: "Table 1: Fall Enrollment and Final Performance Demographics."
#| echo: true

library(knitr)

# Sample compliant data frame
course_data <- data.frame(
  `Course Name` = c("Introduction to Data Science", "Applied Biostatistics"),
  `Enrolled Students` = c(45, 28),
  `Passing Rate` = c("92%", "88%"),
  check.names = FALSE
)

# kable automatically injects accessible HTML table wrappers
knitr::kable(course_data)
```

## Data Visualizations & Alt Text {#sec-visualizations}

Every chart, plot, or image **must** include an explicit alternative text description (`fig-alt`). The text should concisely convey the point of the visualization rather than just stating "a plot".

Additionally, use a colorblind-friendly palette (such as Viridis) to guarantee compliance with WCAG Level AA color contrast ratios (minimum **4.5:1** for normal text, **3:1** for graphical elements).

```{r}
#| label: fig-performance
#| fig-cap: "Figure 1: Distribution of final exam scores across sections."
#| fig-alt: "A histogram displaying a normal distribution of final exam scores ranging from 65 to 100 percent. The distribution peaks sharply around a mean of 84 percent, with a standard deviation of 4 points."
#| echo: true

library(ggplot2)

# Generate dummy normal data
set.seed(42)
scores <- data.frame(grade = rnorm(150, mean = 84, sd = 4))

# Plotting using the colorblind-friendly scale
ggplot(scores, aes(x = grade)) +
  geom_histogram(binwidth = 1.5, fill = "#21918c", color = "#ffffff") +
  theme_minimal(base_size = 14) + # Boost base font size for readability
  labs(
    x = "Final Exam Score (%)",
    y = "Student Count"
  )
```

## Form Accessibility {#sec-activities}

If this document serves as an assignment, ensure form fields or activity spaces are cleanly separated. Provide descriptive instructions instead of color-coded prompts (e.g., avoid "Answer the question highlighted in red").
