---
title: "MLJAR Studio and AutoML"
author: "Eric A. Suess"
date: "December 3, 2026"
date-format: "MMMM D, YYYY"
subtitle: "From assisted notebooks to responsible automated modeling<br>12:15–1:15 PM · Zoom: <https://csueb.zoom.us/j/88674490934><br><small>Slides produced with the help of OpenAI Codex and Pi Agent.</small>"
format:
  revealjs:
    theme: simple
    css: talk05a.css
    slide-number: true
    transition: fade
    incremental: false
    embed-resources: true
    code-fold: true
    code-summary: "Show the code"
    width: 1280
    height: 800
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup-reticulate
#| include: false
matplotlib_cache <- file.path(tempdir(), "matplotlib")
dir.create(matplotlib_cache, showWarnings = FALSE, recursive = TRUE)
Sys.setenv(
  MPLCONFIGDIR = matplotlib_cache,
  MPLBACKEND = "Agg"
)
reticulate::use_python(Sys.which("python3"), required = TRUE)
```

## Today’s goals

By the end of this session, you should be able to:

- Distinguish **MLJAR Studio** from the **MLJAR AutoML** Python package.
- Describe what automated machine learning (**AutoML**) can—and cannot—do.
- Protect an untouched test set while comparing candidate models.
- Connect AutoML to familiar modeling workflows in R and Python.
- Evaluate an AutoML result without treating the highest score as the whole answer.

## Two related—but different—tools

| Tool type | Main purpose | Example |
|---|---|---|
| Notebook workspace | Mix narrative, code, output, and AI assistance | MLJAR Studio |
| AutoML system | Train and compare predictive models | MLJAR AutoML |

**MLJAR Studio** crosses categories: it is a local, notebook-based AI workspace that can run AutoML experiments.

## What is MLJAR Studio? {.compact}

[MLJAR Studio](https://mljar.com/) is a local, Python notebook workspace with AI-assisted data analysis.

- Ask questions in natural language and receive visible Python code.
- Run the generated analysis on the local computer.
- Inspect, edit, and rerun notebook code.
- Use **AutoLab Experiments** for iterative model experiments.
- Use the MLJAR AutoML package from the same workspace.
- Convert suitable notebook results into applications with Mercury.

::: {.callout-important}
Local code execution does not automatically make every AI interaction private. Check the configured model provider and institutional data policy before using sensitive data.
:::

## MLJAR Studio vs. MLJAR AutoML

| MLJAR Studio | `mljar-supervised` / MLJAR AutoML |
|---|---|
| Desktop data science workspace | Open-source Python package |
| Notebook and AI interface | Classification and regression automation |
| Helps generate and run Python | Preprocesses, trains, tunes, and compares models |
| Can launch experiments | Produces model documentation and diagnostics |

You can use the AutoML package inside MLJAR Studio, Positron, VS Code, JupyterLab, or another Python environment.

## What is automated machine learning?

For **supervised learning**, we have predictors $X$ and an outcome $Y$.

- **Classification:** predict a category, such as “spam” or “not spam.”
- **Regression:** predict a numeric value, such as sales or travel time.

AutoML automates repeated parts of the predictive-modeling workflow:

**preprocess** → **fit candidates** → **tune** → **validate** → **compare** → **document**

It searches a defined space. It does not remove the need to define a valid problem.

## What AutoML can automate {.compact}

Depending on the tool and configuration, AutoML may:

- Impute missing values and encode categorical predictors.
- Establish a simple baseline.
- Train linear, tree-based, boosting, neural-network, or neighbor models.
- Tune hyperparameters.
- Use resampling or cross-validation.
- Build ensembles or stacked models.
- Rank models using a selected metric.
- Produce learning curves, importance measures, or SHAP explanations.
- Save models and generate reports.

Automation saves repetition; it does not guarantee a trustworthy model.

## What AutoML cannot decide for you

AutoML does not determine whether:

- The target variable represents the real question.
- The sample represents the population where the model will be used.
- A predictor leaks future or outcome information.
- The metric reflects the cost of mistakes.
- Group performance is equitable.
- Prediction should be interpreted as causation.
- Deployment is ethical, legal, secure, or useful.

These are human and institutional responsibilities.

## Protect the final test set

A defensible predictive workflow separates three roles:

| Data portion | Purpose |
|---|---|
| Training data | Estimate model parameters |
| Validation or resamples | Compare models and tune settings |
| Test data | Estimate final performance on untouched observations |

Repeatedly checking the test set turns it into another validation set.

::: {.callout-warning}
Split by **time, person, school, patient, or site** when random row splitting would leak related information across sets.
:::

## MLJAR AutoML modes

MLJAR provides modes with different defaults:

- **Explain:** faster exploration, interpretable models, and detailed explanations
- **Perform:** balance predictive quality, training time, and prediction speed
- **Compete:** broader search, feature engineering, ensembles, and stacking
- **Optuna:** more intensive hyperparameter optimization

Begin with **Explain** to understand the dataset and baseline. More computation is not automatically more scientific.

<small>Source: [MLJAR AutoML](https://mljar.com/automl/) and [AutoML Modes](https://supervised.mljar.com/features/modes/)</small>

## MLJAR AutoML: a small Python example

Install the package in the intended Python environment:

```bash
python -m pip install mljar-supervised
```

```{python}
#| eval: false
from sklearn.datasets import load_iris
from sklearn.model_selection import train_test_split
from sklearn.metrics import accuracy_score
from supervised import AutoML

data = load_iris(as_frame=True)
X_train, X_test, y_train, y_test = train_test_split(
    data.data, data.target,
    test_size=0.20, stratify=data.target, random_state=310
)

automl = AutoML(
    mode="Explain",
    total_time_limit=300,
    results_path="iris_automl"
)
automl.fit(X_train, y_train)
predictions = automl.predict(X_test)
accuracy_score(y_test, predictions)
```

## Do not stop at the leaderboard {.compact}

After training, inspect:

- The **baseline** and whether complex models improve on it
- The validation method and evaluation metric
- Variation across folds—not only the mean score
- Confusion matrices or residual plots
- Learning curves and signs of overfitting
- Feature importance and explanation limitations
- Errors on meaningful subgroups
- Training time, prediction time, and model complexity
- Performance on the untouched test set

The winning model may not be the model you should deploy.

## The same comparison logic in R

This teaching example manually compares four model families with five-fold cross-validation. AutoML systems automate and expand this pattern.

```{r}
#| label: fig-model-comparison
#| code-fold: true
#| code-summary: "Show the tidymodels code"
#| fig-width: 9
#| fig-height: 4.6
#| fig-align: center
#| out-width: "84%"
#| fig-alt: "Bar chart comparing mean five-fold cross-validation accuracy for four classification models fitted to the iris training data."
library(tidymodels)

set.seed(310)

iris_data <- as_tibble(iris)
iris_split <- initial_split(iris_data, prop = 0.80, strata = Species)
iris_train <- training(iris_split)
iris_folds <- vfold_cv(iris_train, v = 5, strata = Species)

model_specs <- list(
  "Multinomial regression" = multinom_reg() |>
    set_engine("nnet"),
  "Decision tree" = decision_tree() |>
    set_mode("classification") |>
    set_engine("rpart"),
  "Random forest" = rand_forest(trees = 500) |>
    set_mode("classification") |>
    set_engine("ranger"),
  "Nearest neighbors" = nearest_neighbor(neighbors = 5) |>
    set_mode("classification") |>
    set_engine("kknn")
)

model_results <- purrr::imap_dfr(
  model_specs,
  \(spec, model_name) {
    workflow() |>
      add_formula(Species ~ .) |>
      add_model(spec) |>
      fit_resamples(
        resamples = iris_folds,
        metrics = metric_set(accuracy)
      ) |>
      collect_metrics() |>
      transmute(
        model = model_name,
        mean_accuracy = mean,
        std_error = std_err
      )
  }
)

model_results |>
  mutate(model = forcats::fct_reorder(model, mean_accuracy)) |>
  ggplot(aes(x = mean_accuracy, y = model)) +
  geom_col(fill = "#1f77b4", width = 0.7) +
  geom_errorbar(
    aes(
      xmin = pmax(0, mean_accuracy - std_error),
      xmax = pmin(1, mean_accuracy + std_error)
    ),
    width = 0.18,
    color = "#163a5f"
  ) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title = "Candidate Models on the Iris Training Data",
    subtitle = "Mean five-fold cross-validation accuracy ± one standard error",
    x = "Cross-validation accuracy",
    y = NULL,
    caption = "Teaching example: model specifications were chosen manually"
  ) +
  theme_minimal(base_size = 15)
```

## tidymodels and scikit-learn: same workflow

| Modeling task | R with tidymodels | Python with scikit-learn |
|---|---|---|
| Train/test split | `initial_split()` | `train_test_split()` |
| Stratified folds | `vfold_cv()` | `StratifiedKFold()` |
| Preprocessing + model | `workflow()` | `Pipeline` / `make_pipeline()` |
| Resampling | `fit_resamples()` | `cross_val_score()` |
| Accuracy | `yardstick::accuracy()` | `scoring="accuracy"` |
| Final model | `last_fit()` | `fit()` and test-set scoring |

The syntax differs, but the statistical logic is the same. Keep preprocessing inside the workflow or pipeline so it is learned separately within each training fold.

## The same comparison logic in Python

This scikit-learn example compares the same four model families using stratified five-fold cross-validation on the Iris training data.

```{python}
#| label: fig-model-comparison-python
#| code-fold: true
#| code-summary: "Show the scikit-learn code"
#| fig-width: 9
#| fig-height: 4.6
#| fig-align: center
#| out-width: "84%"
#| fig-alt: "Bar chart comparing mean stratified five-fold cross-validation accuracy for four scikit-learn classification models fitted to the Iris training data."
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.datasets import load_iris
from sklearn.ensemble import RandomForestClassifier
from sklearn.linear_model import LogisticRegression
from sklearn.model_selection import (
    StratifiedKFold,
    cross_val_score,
    train_test_split,
)
from sklearn.neighbors import KNeighborsClassifier
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import StandardScaler
from sklearn.tree import DecisionTreeClassifier

iris_data = load_iris(as_frame=True)
X_train, X_test, y_train, y_test = train_test_split(
    iris_data.data,
    iris_data.target,
    test_size=0.20,
    stratify=iris_data.target,
    random_state=310,
)

iris_folds = StratifiedKFold(
    n_splits=5,
    shuffle=True,
    random_state=310,
)

model_specs = {
    "Multinomial logistic regression": make_pipeline(
        StandardScaler(),
        LogisticRegression(max_iter=1000),
    ),
    "Decision tree": DecisionTreeClassifier(random_state=310),
    "Random forest": RandomForestClassifier(
        n_estimators=500,
        random_state=310,
    ),
    "Nearest neighbors": make_pipeline(
        StandardScaler(),
        KNeighborsClassifier(n_neighbors=5),
    ),
}

model_results = []
for model_name, model in model_specs.items():
    fold_accuracy = cross_val_score(
        model,
        X_train,
        y_train,
        cv=iris_folds,
        scoring="accuracy",
    )
    model_results.append(
        {
            "model": model_name,
            "mean_accuracy": fold_accuracy.mean(),
            "std_error": (
                fold_accuracy.std(ddof=1) /
                np.sqrt(len(fold_accuracy))
            ),
        }
    )

model_results = (
    pd.DataFrame(model_results)
    .sort_values("mean_accuracy")
)

fig, ax = plt.subplots(figsize=(9, 4.6))
ax.barh(
    model_results["model"],
    model_results["mean_accuracy"],
    xerr=model_results["std_error"],
    color="#f28e2b",
    ecolor="#7a3e00",
    capsize=4,
)
_ = ax.set_xlim(0, 1)
_ = ax.set_xlabel("Cross-validation accuracy")
_ = ax.set_title(
    "Candidate Models on the Iris Training Data\n"
    "Mean stratified five-fold accuracy ± one standard error",
    fontweight="bold",
    fontsize=12,
    pad=10,
)
_ = ax.grid(axis="x", alpha=0.25)
_ = ax.spines[["top", "right"]].set_visible(False)
_ = fig.tight_layout()
plt.show()
```

## What the teaching examples leave out

The comparisons are reproducible, but they are deliberately small:

- Only four manually selected specifications
- Almost no hyperparameter tuning
- One split and five folds
- Accuracy as the only metric
- A tiny, clean teaching dataset
- No subgroup, fairness, calibration, or deployment analysis

AutoML can search more broadly. Sound statistical reasoning determines whether that search answers the right question.

## A responsible AutoML checklist {.compact}

Before trusting a result, document:

1. **Question:** What decision will the prediction support?
2. **Outcome:** How was the target defined and measured?
3. **Data:** Who or what is represented—and missing?
4. **Split:** Could information leak across training and evaluation?
5. **Metric:** What errors matter most?
6. **Baseline:** Does the model beat a simple rule?
7. **Uncertainty:** How stable is performance?
8. **Equity:** How does performance vary across relevant groups?
9. **Reproducibility:** Can the environment and experiment be rerun?
10. **Oversight:** Who reviews, approves, monitors, and retires the model?

## Mini-activity: evaluate a modeling workflow

A researcher wants to compare tabular prediction models and inspect reports.

Explain:

1. Whether MLJAR Studio, the MLJAR AutoML package, or both fit the task.
2. Which AutoML mode would be a reasonable starting point.
3. How training, validation, and test data should be separated.
4. What evidence beyond the top leaderboard score should be reviewed.

There may be more than one defensible workflow. State the tradeoff.

## Key takeaways {.compact}

- MLJAR Studio is a local notebook and AI workspace; MLJAR AutoML is a Python package.
- AutoML automates model search—not problem formulation, data quality, or accountability.
- Protect the test set, inspect the baseline, and evaluate more than the top score.
- Automated comparison expands a modeling workflow; it does not replace statistical reasoning.

## References {.smaller}

All links were accessed October 3, 2026.

- MLJAR: [MLJAR Studio](https://mljar.com/), [Studio Documentation](https://mljar.com/docs/), [MLJAR AutoML](https://mljar.com/automl/), and [`mljar-supervised` Documentation](https://supervised.mljar.com/)
- R modeling: [tidymodels](https://www.tidymodels.org/), [workflows](https://tidymodels.github.io/workflows/), [rsample](https://rsample.tidymodels.org/), and [yardstick](https://yardstick.tidymodels.org/)
- Python modeling: [scikit-learn User Guide](https://scikit-learn.org/stable/user_guide.html), [Cross-validation](https://scikit-learn.org/stable/modules/cross_validation.html), and [Pipelines](https://scikit-learn.org/stable/modules/compose.html#pipeline)

Product features, licensing, AI providers, and institutional availability can change. Verify current documentation before installation or adoption.

## Questions & discussion

::: {.callout-note}
### Join the Discussion
**Date:** December 3, 2026  
**Time:** 12:15–1:15 PM  
**Zoom:** <https://csueb.zoom.us/j/88674490934>
:::

- Where would MLJAR Studio or the MLJAR AutoML package fit into your current workflow?
- What would you need to verify before trusting an AutoML result?
