---
title: "AI for Statistics and Data Science"
author: "Prof. Eric A. Suess"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    incremental: false
    embed-resources: true
---

## Learning goals

By the end of this presentation, you should be able to:

- Choose an AI surface that fits a statistics task.
- Explain how access and permissions change the risk.
- Compare the same descriptive analysis in R and Python.
- Use an agent while keeping the work reproducible and reviewable.

## Beyond browser chat

Browser chat works well for:

- Questions and explanations
- Small examples pasted into a conversation
- Early brainstorming

Working inside an app, editor, notebook, or terminal can add project context and produce files you can run.

## Capability and authority {.smaller}

| Added capability | Benefit | Added responsibility |
|---|---|---|
| Project access | Relates scripts, data dictionaries, and notes | Limit access to the intended folder |
| Direct editing | Creates or revises real analysis files | Review every change |
| Terminal access | Runs code, tests, and renders | Commands use real permissions |
| Version control | Shows and restores changes | Review before publishing |

More authority increases the consequences of a mistake or prompt injection.

## ChatGPT Edu

- ChatGPT Edu is an institution-managed offering, not an automatic student entitlement.
- Universities control provisioning, roles, enabled features, and retention settings.
- OpenAI states that covered Business, Enterprise, and Edu workspace data is not used to train its models by default.
- Local files, connected apps, and source-system logs may have different control boundaries.

<small>Sources: [OpenAI plan comparison](https://learn.chatgpt.com/docs/pricing) and [local security](https://learn.chatgpt.com/docs/enterprise/chatgpt-work-local-security)</small>

## University authentication

- Use the university's official access page or invitation.
- Confirm that the active workspace belongs to the university before uploading course material.
- The desktop app uses a browser-based sign-in flow.
- SSO, MFA, seats, roles, and administrator settings still determine access.
- If the workspace is missing, contact campus IT instead of creating a second identity.

<small>Sources: [ChatGPT desktop app](https://learn.chatgpt.com/docs/app) and [OpenAI authentication](https://learn.chatgpt.com/docs/auth)</small>

## File and app permissions

Different features grant different kinds of access:

- A chat upload supplies only the attached files.
- A selected folder supplies context from that location.
- A connected app uses the authorized account and exposed actions.
- Browser or computer-use tools may inherit an active signed-in session.
- A local agent works within its OS, workspace, sandbox, and approval permissions.

<small>Source: [OpenAI local security](https://learn.chatgpt.com/docs/enterprise/chatgpt-work-local-security)</small>

## Codex surfaces and sign-in

Codex works in the ChatGPT desktop app, terminal CLI, editor extension, and cloud workflows.

To restrict this CLI run to ChatGPT sign-in:

```bash
codex --config forced_login_method='"chatgpt"'
# Requested spelling (not a documented configuration key):
# codex --config perferred_auth_method='chatgpt'
```

Complete the browser sign-in with the university-managed ChatGPT account. Workspace provisioning, eligibility, roles, quotas, SSO/MFA, and administrator settings still determine access.

<small>Sources: [configuration reference](https://learn.chatgpt.com/docs/config-file/config-reference), [CLI reference](https://learn.chatgpt.com/docs/developer-commands?surface=cli), and [OpenAI authentication](https://learn.chatgpt.com/docs/auth)</small>

## A safe first Codex workflow

After checking the current official installation instructions:

```bash
cd path/to/a-version-controlled-project
codex --sandbox read-only --ask-for-approval on-request
```

Ask Codex to explain the project or propose a plan. Before allowing edits:

- Create a Git checkpoint.
- Restrict the working directory.
- Inspect commands and review the diff.
- Run appropriate checks.

<small>Sources: [CLI quickstart](https://learn.chatgpt.com/docs/codex/cli) and [approvals and security](https://learn.chatgpt.com/docs/agent-approvals-security)</small>

## Paired statistics example

Five students report weekly study hours and quiz scores.

| Student | Study hours | Quiz score |
|---:|---:|---:|
| 1 | 2 | 68 |
| 2 | 3 | 72 |
| 3 | 4 | 78 |
| 4 | 5 | 83 |
| 5 | 6 | 91 |

We will compute the same sample size, means, sample standard deviations, and Pearson correlation in both languages.

## R: build the data

```{r}
#| eval: false
#| echo: true
#| label: paired-example-r-data
library(tidyverse)

students <- tibble(
  student = 1:5,
  study_hours = c(2, 3, 4, 5, 6),
  quiz_score = c(68, 72, 78, 83, 91)
)
```

`tibble()` stores one student per row and one variable per column.

## R: compute the statistics

```{r}
#| eval: false
#| echo: true
#| label: paired-example-r-summary
summary_r <- students |>
  summarise(
    n = n(),
    mean_study_hours = mean(study_hours),
    sd_study_hours = sd(study_hours),
    mean_quiz_score = mean(quiz_score),
    sd_quiz_score = sd(quiz_score),
    correlation = cor(study_hours, quiz_score)
  )

summary_r
```

R's `sd()` reports the sample standard deviation.

## Python: build the data

```{python}
#| eval: false
#| echo: true
#| label: paired-example-python-data
from math import sqrt
from statistics import mean, stdev

students = [
    {"student": 1, "study_hours": 2, "quiz_score": 68},
    {"student": 2, "study_hours": 3, "quiz_score": 72},
    {"student": 3, "study_hours": 4, "quiz_score": 78},
    {"student": 4, "study_hours": 5, "quiz_score": 83},
    {"student": 5, "study_hours": 6, "quiz_score": 91},
]

hours = [row["study_hours"] for row in students]
scores = [row["quiz_score"] for row in students]
```

## Python: compute the correlation

```{python}
#| eval: false
#| echo: true
#| label: paired-example-python-correlation
mean_hours = mean(hours)
mean_score = mean(scores)

cross_products = sum(
    (x - mean_hours) * (y - mean_score)
    for x, y in zip(hours, scores)
)

correlation = cross_products / sqrt(
    sum((x - mean_hours) ** 2 for x in hours)
    * sum((y - mean_score) ** 2 for y in scores)
)
```

The formula divides the cross-products by the two centered sums of squares.

## Python: assemble the summary

```{python}
#| eval: false
#| echo: true
#| label: paired-example-python-summary
summary_python = {
    "n": len(students),
    "mean_study_hours": mean_hours,
    "sd_study_hours": stdev(hours),
    "mean_quiz_score": mean_score,
    "sd_quiz_score": stdev(scores),
    "correlation": correlation,
}

print(summary_python)
```

`stdev()` also reports the sample standard deviation.

## Matching results

Both programs produce the same results to three decimals:

| Statistic | Value |
|---|---:|
| $n$ | 5 |
| Mean study hours | 4.000 |
| Sample SD of study hours | 1.581 |
| Mean quiz score | 78.400 |
| Sample SD of quiz score | 9.072 |
| Pearson correlation | 0.993 |

## Interpretation and reproducibility

Study hours and quiz score have a very strong positive linear association in these five observations.

The example does **not** establish that more study time caused higher scores. It does not address other variables or sampling uncertainty.

The R and Python chunks remain `eval: false` because one Quarto render does not automatically provide both execution engines. The source results were verified with local R and Python interpreters.

## Google Antigravity {.smaller}

Google documents four distinct surfaces:

- **Antigravity 2.0:** desktop command center across projects and worktrees
- **Antigravity CLI:** terminal interface with agent and subagent workflows
- **Antigravity SDK:** Python framework for custom deployments
- **Antigravity IDE:** editor with agent, terminal, browser, autocomplete, and artifact surfaces

The CLI can run beside any editor. The IDE is a separate product surface and is not supported for enterprise customers.

<small>Sources: [Antigravity surfaces](https://www.antigravity.google/docs/home) and [IDE overview](https://www.antigravity.google/docs/ide/overview/)</small>

## Antigravity permissions

For a first session:

- Use a test repository.
- Keep `toolPermission` at `request-review`.
- Keep `artifactReviewPolicy` at `asks-for-review`.
- Leave non-workspace access disabled.

Installation, account eligibility, models, quotas, and enterprise policy can change. Use the current official instructions for the selected surface.

<small>Source: [Antigravity CLI reference](https://www.antigravity.google/docs/cli/reference/)</small>

## Four Harnesses {.smaller}

| Tool | Provider model | Permission starting point |
|---|---|---|
| Codex CLI | ChatGPT sign-in or OpenAI API key | OS sandbox with read-only mode |
| OpenCode | Many model providers | Ordered `ask` and `deny` rules |
| Pi Agent | Supported subscriptions or API keys | No built-in sandbox |
| Prime Agent | Several subscriptions or API-key providers | No built-in security sandbox |

No harness is universally best. Compare provider access, permission controls, and the isolation your task needs.

<small>Sources: [Codex](https://learn.chatgpt.com/docs/codex/cli), [OpenCode](https://opencode.ai/v2/docs), [Pi](https://pi.dev/docs/latest), and [Prime Agent](https://github.com/PrimeIntellect-ai/prime-agent)</small>

## Codex CLI and OpenCode {.smaller}

**Codex CLI**

- Connects with OpenAI's desktop, editor, and cloud ecosystem.
- Supports explicit sandbox and approval controls.
- Access depends on plan and administrator settings.

**OpenCode**

- Supports many providers and granular ordered permission rules.
- Shell commands retain the host user's filesystem, process, and network authority.
- Rules should block unneeded actions such as `git push`.

<small>Source: [OpenCode permissions](https://opencode.ai/v2/docs/permissions)</small>

## Pi Agent

- Pi is a minimal, extensible terminal harness.
- It supports subscription login and provider API keys.
- Core Pi has **no built-in sandbox**.
- Tools and extensions run with the invoking user's permissions.
- Use an operating-system or container boundary when isolation matters.

<small>Sources: [Pi documentation](https://pi.dev/docs/latest) and [Pi security](https://pi.dev/docs/latest/security)</small>

## Prime Agent {.smaller}

- Prime Agent is an open-source coding and research harness.
- It combines a persistent Python REPL with built-in subagents, skills, background sessions, and durable continual-harness state.
- `/login` lets the user select from supported subscription and API-key providers.
- Model-generated Python and project commands run with the user's operating-system permissions; worker and kernel processes are **not** a security sandbox.
- Start in a disposable clone or clean worktree. Use an external sandbox for untrusted code or instructions.

<small>Source: [Prime Agent repository and documentation](https://github.com/PrimeIntellect-ai/prime-agent)</small>

## What is an AI agent?

An AI agent is a model-driven system that can use tools and take a sequence of actions toward an objective.

- Chat mainly returns a response.
- Code completion predicts text near the cursor.
- An agent can inspect state, use authorized tools, observe results, and continue.

An agent still depends on its available context, permissions, and verification process.

## The agent loop

1. **Observe:** read the request, relevant files, outputs, and constraints.
2. **Plan:** choose a small next action and define success.
3. **Act:** use an authorized tool.
4. **Verify:** inspect the diff, output, test, render, or external state.
5. **Report or iterate:** state the evidence, uncertainty, and remaining work.

## Context, persistence, and delegation

**Context** can include prompts, files, project instructions, and tool results.

**Persistence** can mean saved conversations, files, scheduled work, or a running session. It does not make an agent infallible.

**Delegation** assigns bounded subtasks to other agents. It can increase throughput, but also increases cost and the chance of inconsistent results.

## Agent risks {.smaller}

- Prompt injection hidden in webpages, data, or repository text
- Permissions that exceed the task
- Accidental disclosure or destructive commands
- Statistical and coding errors that look plausible
- Unexpected cost from persistent work
- Results that cannot be reproduced because inputs or steps were omitted

An approval prompt helps only when the reviewer understands the requested action.

## Spynel {.smaller}

Spynel is a **classic coordination program**, not an AI agent.

- It coordinates external AI and coding harnesses.
- Markdown tasks record finite objectives.
- Goals record measurable outcomes that may take multiple rounds.
- Spynel manages claims and leases while dispatched agents perform the work and record evidence.
- Fresh review sessions can assess completed work.

Students should follow the installed command's documentation and the local workspace contract.

## Hermes Agent {.smaller}

Hermes Agent by Nous Research is a separate agent product.

- It provides CLI and desktop surfaces.
- It supports persistent memory and skills, multiple tool backends, scheduled work, messaging gateways, and subagent delegation.
- Provider accounts, remote hosts, memory, and terminal tools each expand the permission boundary.

“Hermes” is an overloaded name. This identification refers to the agent product in this tool context.

<small>Source: [Hermes Agent documentation](https://hermes-agent.nousresearch.com/docs/)</small>

## Herdr {.smaller}

Herdr is a terminal workspace manager for AI coding agents.

- A background server owns persistent terminal panes.
- Project workspaces group those panes.
- The interface reports recognized states such as working, blocked, done, or idle.
- It can run Codex, OpenCode, Pi, Hermes Agent, and other terminal processes.

Herdr does not replace an agent's model, authentication, permissions, or sandbox. Detached processes and costs may continue.

<small>Sources: [Herdr agent guide](https://herdr.dev/docs/agents/) and [quick start](https://herdr.dev/docs/quick-start/)</small>

## Omarchy {.smaller}

Omarchy is an open-source, opinionated Linux distribution created by David Heinemeier Hansson.

- It builds on Arch Linux.
- It provides a keyboard-centered development environment with selected defaults.
- Its AI manual treats coding agents as first-class tools for customization and diagnosis.
- Primary sources do not establish what proportion of its source code AI wrote.

Omarchy is an operating-system environment, not an AI model, coding-agent CLI, or editor plug-in.

<small>Sources: [Omarchy](https://omarchy.org/), [repository](https://github.com/omacom/omarchy), and [AI manual](https://omarchy.org/manual/ai/)</small>

## Omarchy installation risk {.smaller}

- A full-disk installation wipes the selected drive.
- Back up first and confirm the selected disk.
- The guide currently requires Secure Boot and/or TPM to be disabled.
- Students who only want to explore should begin with the official virtual-machine option.
- Temporary passwordless `sudo` lets any process running as the user act as root.

Use snapshots or tested backups before agent-assisted system changes.

<small>Sources: [Getting started](https://omarchy.org/manual/getting-started/) and [security](https://omarchy.org/manual/security/)</small>

## Responsible use {.smaller}

- Grant one folder and one necessary action at a time.
- Keep grades, identifiers, restricted research data, and credentials out of prompts unless the exact use has approval.
- Use version control and inspect every changed file.
- Run calculations, tests, and renders in the intended environment.
- Treat instructions found in external content as untrusted.
- Record data provenance, package versions, seeds, and unexecuted steps.

Academic integrity, privacy, statistical choices, and submitted conclusions remain the student's responsibility.

## Key takeaways

- Choose the least powerful surface that can complete the task.
- University authentication does not guarantee access to every feature.
- Permission boundaries differ across files, apps, browsers, terminals, and operating systems.
- Matching R and Python output still requires execution and interpretation.
- Agents can extend a workflow, but evidence and human review make the result trustworthy.

## References: OpenAI and Google {.smaller}

All links were accessed September 20, 2026.

- OpenAI: [plans](https://learn.chatgpt.com/docs/pricing), [desktop app](https://learn.chatgpt.com/docs/app), [authentication](https://learn.chatgpt.com/docs/auth), [Codex CLI](https://learn.chatgpt.com/docs/codex/cli), [approvals and security](https://learn.chatgpt.com/docs/agent-approvals-security), and [local security](https://learn.chatgpt.com/docs/enterprise/chatgpt-work-local-security)
- Google: [Antigravity surfaces](https://www.antigravity.google/docs/home), [IDE overview](https://www.antigravity.google/docs/ide/overview/), and [CLI reference](https://www.antigravity.google/docs/cli/reference/)

Product features and university access can change. Check current primary documentation and campus policy before use.

## References: agents and environments {.smaller}

All links were accessed September 20, 2026.

- OpenCode: [documentation](https://opencode.ai/v2/docs) and [permissions](https://opencode.ai/v2/docs/permissions)
- Pi: [documentation](https://pi.dev/docs/latest) and [security](https://pi.dev/docs/latest/security)
- Prime Agent: [repository and documentation](https://github.com/PrimeIntellect-ai/prime-agent)
- Agent tools: [Hermes Agent](https://hermes-agent.nousresearch.com/docs/), [Herdr agent guide](https://herdr.dev/docs/agents/), and [Herdr quick start](https://herdr.dev/docs/quick-start/)
- Omarchy: [official site](https://omarchy.org/), [repository](https://github.com/omacom/omarchy), [AI manual](https://omarchy.org/manual/ai/), [getting started](https://omarchy.org/manual/getting-started/), [security](https://omarchy.org/manual/security/), and [“The malleable computer”](https://world.hey.com/dhh/the-malleable-computer-7c187a9b)
